Non-word repetition in Yélî Dnye

### 2020-12-15 manuscript.Rmd
create trnsc.inc to replace trnsc in main analyses
trnsc maintained two participants we had decided to exclude (c105 and c78)

trnsc$Sex=factor(trnsc$Sex)
trnsc$birthOrder=as.numeric(trnsc$birthOrder)
```

@@

aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$attempt),mean)->pwcm_att
colnames(pwcm_att)<-c("ID","Age","Attempt","Word")

#remove children who do not have both first & subsequent attempts
# which is
excl=sum(table(pwcm_att$ID)!=2)

@@

```{r exclude subsequent}
trnsc=trnsc[trnsc$attempt=="first",]

@@

```{r overall}
aggregate(trnsc$correct*100,by=list(trnsc$id),mean)->overall
aggregate(trnsc$phon_score*100,by=list(trnsc$id),mean)->phon_score
aggregate(trnsc$nld*100,by=list(trnsc$id),mean)->nld
overall=cbind(overall,phon_score$x,nld$x)
overall=merge(overall, demo[,c("ID","Sex","Age","birthOrder","matEd")], by.x="Group.1",by.y="ID", all.x=T)
colnames(overall)<-c("ID","Word","Phoneme","NLD","Sex","Age","BrthO","matEd")

@@

The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`. Scores based on phonemes are even higher  M = `r round(mean(overall$Phoneme))`% (SD = `r round(sd(overall$Phoneme))`%), Cohen's d = `r cohensd(mean(overall$Phoneme),sd(overall$Phoneme))`. The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted. Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.

@@

```{r xling-freq-ana}
trnsc$avg_fr_s=scale(trnsc$avg_fr)
trnsc$age.rounded_s=scale(trnsc$age.rounded)
xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
mycoeff=round(summary(xl_fr)$coefficients,3)
#summary(xl_fr)
#plot(xl_fr)

aggregate(trnsc$correct*100,by=list(trnsc$avg_fr_s),mean)->xl_fr_tab
colnames(xl_fr_tab)<-c("xlf","Word")
xlfr_r=cor.test(xl_fr_tab$xlf,xl_fr_tab$Word)
```

@@

```{r Fig2-xling-freq, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages."}
aggregate(trnsc$correct*100,by=list(trnsc$target,trnsc$avg_fr),mean)->xling
colnames(xling)<-c("Target","Avg_fr","Word")
stims.ipa <- read_delim("stimuli3.txt", delim = "\t") %>%
  select(target, target_ortho, phono)
xling <- merge(xling, stims.ipa, by.x = "Target", by.y = "target")

@@

```{r cor vs xl}
#fr_cor=glmer(correct~avg_fr_cor+age.rounded+(1|file) + (1|target),data=trnsc,family="binomial")
#fr is not sig, but close (p=.1, estimate is 17)
#trnsc$avg_fr_cor_s=scale(trnsc$avg_fr_cor)
#both_fr=glmer(correct~avg_fr_s+avg_fr_cor_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
#both_coeff=round(summary(both_fr)$coefficients,3)

#fr_cor_ty=glmer(correct~avg_fr_cor_ty+age.rounded+(1|file) + (1|target),data=trnsc,family="binomial")
#fr is sig
trnsc$avg_fr_cor_ty_s=scale(trnsc$avg_fr_cor_ty)
both_fr=glmer(correct~avg_fr_s+avg_fr_cor_ty_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
both_coeff=round(summary(both_fr)$coefficients,3)

aggregate(trnsc$correct*100,by=list(trnsc$avg_fr_cor_ty_s),mean)->cor_fr_tab
colnames(cor_fr_tab)<-c("corf","Word")

@@

#remove English meanings
trnsc$meaning[grep("English",trnsc$meaning)]<-NA

@@

Finally, we looked into the frequency with which mispronunciations resulted in real words. Two thirds of them were: `r round(sum(!is.na(trnsc$meaning) & trnsc$meaning != "0")/sum(!is.na(trnsc$mispronunciation))*100)`%. This is a type of analysis that is seldom reported, and we could only find one comparison point: @castro1998 found that illiterate adults NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate adults did so in only `r round(2/117*100,2)`%. The percentage we observe here is much higher than this, but we do not know whether age, language, or even test structure explains this difference.

@@

aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$type),mean)->by_length
aggregate(trnsc$phon_score*100,by=list(trnsc$id,trnsc$type),mean)->phon_score
aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$type),mean)->nld
by_length=cbind(by_length,phon_score$x,nld$x)
by_length=merge(by_length, demo[,c("ID","Sex","Age","birthOrder","matEd")], by.x="Group.1",by.y="ID", all.x=T)
colnames(by_length)<-c("ID","Length","Word","Phoneme","NLD","Sex","Age","BrthO","matEd")

aggregate(trnsc$correct*100,by=list(trnsc$type),mean)->by_length_no_child
colnames(by_length_no_child)<-c("Length","Word")
by_length_no_child=by_length_no_child[-1,]
length_r=cor.test(by_length_no_child$Length,by_length_no_child$Word)

len_lm=glmer(correct~type+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial",subset=c(trnsc$type!=1))
mycoeff=round(summary(len_lm)$coefficients,3)

>>>>

trnsc$Sex=factor(trnsc$Sex)
trnsc$birthOrder=as.numeric(trnsc$birthOrder)

# Remove excluded children from trnsc
trnsc[which(trnsc$id %in% inc$ID),]->trnsc.inc

@@

aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$age.rounded,trnsc.inc$attempt),mean)->pwcm_att
colnames(pwcm_att)<-c("ID","Age","Attempt","Word")

#remove children who do not have both first & subsequent attempts
# which is
excl=sum(table(pwcm_att$ID)==1)

@@

```{r exclude subsequent}
trnsc.inc=trnsc.inc[trnsc.inc$attempt=="first",]

@@

```{r overall}
aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id),mean)->overall
aggregate(trnsc.inc$phon_score*100,by=list(trnsc.inc$id),mean)->phon_score
aggregate(trnsc.inc$nld*100,by=list(trnsc.inc$id),mean)->nld
overall=cbind(overall,phon_score$x,nld$x)
overall=merge(overall, demo[,c("ID","Sex","Age","birthOrder","matEd")], by.x="Group.1",by.y="ID", all.x=T)
colnames(overall)<-c("ID","Word","Phoneme","NLD","Sex","Age","BrthO","matEd")

@@

The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`. Scores based on phonemes are even higher  M = `r round(mean(overall$Phoneme))`% (SD = `r round(sd(overall$Phoneme))`%), Cohen's d = `r cohensd(mean(overall$Phoneme),sd(overall$Phoneme))`. The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted (the normalized Levenshtein distance is the complement of phoneme-based scores).

@@

```{r xling-freq-ana}
trnsc.inc$avg_fr_s=scale(trnsc.inc$avg_fr)
trnsc.inc$age.rounded_s=scale(trnsc.inc$age.rounded)
xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial")
mycoeff=round(summary(xl_fr)$coefficients,3)
#summary(xl_fr)
#plot(xl_fr)

aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$avg_fr_s),mean)->xl_fr_tab
colnames(xl_fr_tab)<-c("xlf","Word")
xlfr_r=cor.test(xl_fr_tab$xlf,xl_fr_tab$Word)
```

@@

```{r Fig2-xling-freq, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages."}
aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$target,trnsc.inc$avg_fr),mean)->xling
colnames(xling)<-c("Target","Avg_fr","Word")
stims.ipa <- read_delim("stimuli3.txt", delim = "\t") %>%
  select(target, target_ortho, phono)
xling <- merge(xling, stims.ipa, by.x = "Target", by.y = "target")

@@

```{r cor vs xl}
#fr_cor=glmer(correct~avg_fr_cor+age.rounded+(1|file) + (1|target),data=trnsc.inc,family="binomial")
#fr is not sig, but close (p=.1, estimate is 17)
#trnsc.inc$avg_fr_cor_s=scale(trnsc.inc$avg_fr_cor)
#both_fr=glmer(correct~avg_fr_s+avg_fr_cor_s+age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial")
#both_coeff=round(summary(both_fr)$coefficients,3)

#fr_cor_ty=glmer(correct~avg_fr_cor_ty+age.rounded+(1|file) + (1|target),data=trnsc.inc,family="binomial")
#fr is sig
trnsc.inc$avg_fr_cor_ty_s=scale(trnsc.inc$avg_fr_cor_ty)
both_fr=glmer(correct~avg_fr_s+avg_fr_cor_ty_s+age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial")
both_coeff=round(summary(both_fr)$coefficients,3)

aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$avg_fr_cor_ty_s),mean)->cor_fr_tab
colnames(cor_fr_tab)<-c("corf","Word")

@@

#remove English meanings
trnsc.inc$meaning[grep("English",trnsc.inc$meaning)]<-NA

@@

Finally, we looked into the frequency with which mispronunciations resulted in real words. Two thirds of them were: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%. This is a type of analysis that is seldom reported, and we could only find one comparison point: @castro1998 found that illiterate adults NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate adults did so in only `r round(2/117*100,2)`%. The percentage we observe here is much higher than this, but we do not know whether age, language, or even test structure explains this difference.

@@

aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$type),mean)->by_length
aggregate(trnsc.inc$phon_score*100,by=list(trnsc.inc$id,trnsc.inc$type),mean)->phon_score
aggregate(trnsc.inc$nld*100,by=list(trnsc.inc$id,trnsc.inc$type),mean)->nld
by_length=cbind(by_length,phon_score$x,nld$x)
by_length=merge(by_length, demo[,c("ID","Sex","Age","birthOrder","matEd")], by.x="Group.1",by.y="ID", all.x=T)
colnames(by_length)<-c("ID","Length","Word","Phoneme","NLD","Sex","Age","BrthO","matEd")

aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$type),mean)->by_length_no_child
colnames(by_length_no_child)<-c("Length","Word")
by_length_no_child=by_length_no_child[-1,]
length_r=cor.test(by_length_no_child$Length,by_length_no_child$Word)

len_lm=glmer(correct~type+age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial",subset=c(trnsc.inc$type!=1))
mycoeff=round(summary(len_lm)$coefficients,3)
